---
title: "Demographics"
author: "Nate Lant"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nhts2017)
library(tidyverse)
library(plotly)
library(pander)
library(ggpubr)
library(Hmisc)
library(DT)

source("R/output_functions.R")

nhts_persons <- nhts2017::nhts_persons
```

## What percent of the U.S. population uses a wheelchair?
This study compares the mobility of the abled population, the disabled population, and the population that uses wheelchairs. To make this comparison, I created a new variable called "Ability", where each person is labelled "Abled", "Disabled", or "Wheelchair".

In the 2017 NHTS, the respondents were asked if they have a medical condition or handicap that makes it difficult to travel outside of the home. If they responded yes, they were asked if they have had this condition for less than six months, more than six months or their whole life. The next follow-up question asked them to select all that apply from a list of medical devices they use. This list included a cane, walker, seeing eye-dog, crutches, motorized scooter, manual sheelchair, motorized wheelchair, or something else.

Those that responded disabled for more than six months or their whole life were labelled "Disabled". Those that responded disabled (for more than six months or their whole life) and used a wheelchair (either manual or mechanical) were labelled "Wheelchair" (note the "Disabled" population does not include wheelchair users). The rest were labelled "Abled".

```{r ability}
# install.packages("devtools")
# devtools::install_github("byu-transpolab/nhts2017")

# library(nhts2017)
# library(tidyverse)
# library(pander)
# library(scales)

# Create the "Ability" variable in nhts_persons.
nhts_persons <- nhts_persons %>% 
  left_join(nhts_households %>% select(houseid, hhfaminc), 
            by = "houseid") %>%
  mutate(Ability = case_when(w_chair == "07" | w_mtrchr == "08" ~ "Wheelchair",
                             medcond6 == "02" | medcond6 == "03" ~ "Disabled",  
                             medcond == "02" ~ "Abled")) %>% 
  
  filter(Ability != "NA")

# I also want to group age into bin size of 10 years and income into three groups, "low", "med", and "high"

```

The NHTS uses weights on each variable to estimate the total population. In my comparisons, I will show the number of survey respondents, the weighted population, and the distribution of the population (I also filtered out NA responses, as they account for 0.6% of the population).

```{r percent}
nhts_persons %>% group_by(Ability) %>%
  summarise(Survey = n(),
            population = sum(wtperfin)) %>%
  mutate(Population = population,
         `Distribution(%)` = round(population/sum(population)*100, 1)) %>%
  select(-population) %>%
  pander()
```


### Age

The age demographic of wheelchair users is obviously different from the able bodied population, according to the bar plot below. For this study, the population ages 18-64 will be considered. Perhaps in the future, more specific research could be done on the older population, where on-demand transportation could better serve the healthcare industry. 

**Summary Statistics of Age**
```{r age.summary}
# Summary of all ages
nhts_persons %>% group_by(Ability) %>%
  summarise(Mean = mean(r_age),
            Median = median(r_age)) %>%
  pander()
```

```{r age.distribution plot}
nhts_persons <- nhts_persons %>% 
  mutate(Age = case_when(r_age >= 00 & r_age < 10 ~ "0 - 10",
                         r_age >= 10 & r_age < 20 ~ "10 - 20",
                         r_age >= 20 & r_age < 30 ~ "20 - 30",
                         r_age >= 30 & r_age < 40 ~ "30 - 40",
                         r_age >= 40 & r_age < 50 ~ "40 - 50",
                         r_age >= 50 & r_age < 60 ~ "50 - 60",
                         r_age >= 60 & r_age < 70 ~ "60 - 70",
                         r_age >= 70 & r_age < 80 ~ "70 - 80",
                         r_age >= 80 & r_age < 90 ~ "80 - 90",
                         r_age >= 90 & r_age < 100 ~ "90 - 100"))  

nhts_persons %>% group_by(Ability, Age) %>%
  summarise(population = sum(wtperfin)) %>%
  mutate(Distribution = population/sum(population)) %>%
  ggplot(aes(x = Age, y = Distribution)) + 
    geom_col(aes(fill = Ability), colour = "Black", position = "dodge") +
    ggtitle("Distribution of Age", "Comparison of Ability") +
    labs(x = "Age", y = "Distribution") +
    scale_fill_brewer(palette = "PuBuGn", direction = 2) + 
    theme(axis.text.x = element_text(size  = 10, 
                                     angle = 45,
                                     hjust = 1,
                                     vjust = 1))
```



**Summary Statistics of Ages 18-64**
```{r age.summary.filtered}
# Update the nhts_persons variable
nhts_persons <- nhts_persons %>% filter(r_age >= 18 & r_age <= 64)

# Summary of ages 18-64
nhts_persons %>% group_by(Ability) %>%
  summarise(Mean = mean(r_age),
            Median = median(r_age)) %>%
  pander()
```


### Income
The trends in income are very different! This raises questions about how mobility patterns might be more variant because of income rather than because of disabilities.


```{r income.distribution}
nhts_persons %>% group_by(Ability, hhfaminc) %>%
    filter(hhfaminc > 0) %>%
    summarise(Population = sum(wtperfin)) %>%
    as_factor() %>%
    mutate(Distribution = Population/sum(Population)) %>%
  ggplot(aes(x = hhfaminc, y = Distribution)) + 
    geom_col(aes(fill = Ability), colour = "Black", position = "dodge") +
    ggtitle("Distribution of Income Ages 18-64", "Comparison of Ability") +
    labs(x = "Income", y = "Distribution") +
    scale_fill_brewer(palette = "PuBuGn", direction = 2) + 
    theme(axis.text.x = element_text(size  = 10, 
                                     angle = 45,
                                     hjust = 1,
                                     vjust = 1))
```

The NHTS reported income in the categories shown above, not as a quantitative variable. In the following pages, I want to highlight how income affects mobility, so income will be divided into the following groups for convenience.


```{r income.divide}
# Create income group varaible "Income"
nhts_persons <- nhts_persons %>%
  mutate(Income = case_when(hhfaminc == "01" | hhfaminc == "02" | hhfaminc == "03" ~ "Low",
                            hhfaminc == "04" | hhfaminc == "05" ~ "Mid-Low",
                            hhfaminc == "06" | hhfaminc == "07" ~ "Mid-High",
                            hhfaminc == "08" | hhfaminc == "09" | hhfaminc == "10" | hhfaminc == "11" ~ "High"),
         Income = fct_relevel(Income, "Low", "Mid-Low", "Mid-High", "High")) %>%
  filter(Income != "NA")
```

 + *Low* = Less than $25,000
 + *Mid-Low* = \$25,000 - \$50,000
 + *Mid-High* = \$50,000 - \$100,000
 + *High* = More than $100,000
 
```{r income.group.distribution}
# Distribution plot
nhts_persons %>% group_by(Ability, Income) %>%
  summarise(Population = sum(wtperfin)) %>% 
    mutate(Distribution = Population/sum(Population)) %>%
  ggplot(aes(x = Income, y = Distribution)) + 
    geom_col(aes(fill = Ability), colour = "Black", position = "dodge") +
    ggtitle("Distribution of Income by Groups Ages 18 - 64", "Comparison of Ability") +
    labs(x = "Income", y = "Distribution") +
    scale_fill_brewer(palette = "PuBuGn", direction = 2) + 
    theme(axis.text.x = element_text(size  = 10, 
                                     angle = 45,
                                     hjust = 1,
                                     vjust = 1))
```

The interactive table below shows the number of individuals that responded to the survey, their income group, and their ability. The NHTS uses weights to calculate the total population based on the number of respondents, then the table shows the distribution of that weighted population for each subgroup.
```{r income.stats}
nhts_persons %>% group_by(Ability, Income) %>%
  summarise(Survey = n(),
            population = round(sum(wtperfin), 0)) %>%
  mutate(Population = population,
         `Distribution(%)` = round(population/sum(population)*100, 1)) %>%
  select(-population) %>%
  datatable()
```

### Employment Status
The NHTS asked respondents if they work, and those that respond yes were asked if they work full-time or part-time. The results are shown below and are extremely contrast. Is unemployment at all related to limited access to transportation?
```{r employment}
nhts_persons <- nhts_persons %>%
  mutate(Worker = case_when(wkftpt == "01" ~ "Full-Time",
                            wkftpt == "02" ~ "Part-Time",
                            wkftpt == "-1" ~ "Unemployed")) %>%
  filter(Worker != "NA")

nhts_persons %>% group_by(Ability, Worker) %>%
  summarise(Population = sum(wtperfin)) %>% 
    mutate(Distribution = Population/sum(Population)) %>%
  ggplot(aes(x = Worker, y = Distribution)) + 
    geom_col(aes(fill = Ability), colour = "Black", position = "dodge") +
    ggtitle("Distribution of Worker Status Groups Ages 18 - 64", "Comparison of Ability") +
    labs(x = "Worker Status", y = "Distribution") +
    scale_fill_brewer(palette = "PuBuGn", direction = 2) + 
    theme(axis.text.x = element_text(size  = 10, 
                                     angle = 45,
                                     hjust = 1,
                                     vjust = 1))
```
 
```{r store.data}
## Storing the persons file to use in other .Rmd
# 1. Joins households with houseid and hhfaminc
# 2. Creates Ability variable (and filters "NA")
# 3. Creates age groups
# 4. Filters age to 18-64
# 5. Creates Income variable, reorders the levels and filters "NA"
nhts_persons %>%
  mutate(hhpersonid = paste(houseid, personid, sep = "-")) %>% 
  write_rds("data/edited_persons.rds") 
# 6. Creates the "houseid-personid" unique id for each person
```
 
 