---
title: "Trip Purpose"
author: "Nate Lant"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(scales) # need for the axis as percent

library(nhts2017)
library(tidyverse)
library(plotly)
library(pander)
library(ggpubr)
library(Hmisc)

```

```{r persons.data, message=F, warning=F}
# Creating variables and filtering the persons population
persons_edited <- read_rds("data/edited_persons.rds") 

# This data will contain all trips (all households, persons, etc.)
# First find all trips that belong to an hhpersonid, group them by purpose and count them
new_trips <- nhts_trips %>% 
  mutate(hhpersonid = paste(houseid, personid, sep = "-")) %>%
  filter(hhpersonid %in% persons_edited$hhpersonid) %>%
  group_by(hhpersonid, trippurp) %>%
  tally() %>%
  mutate(n = ifelse(n > 10, 10, n))


# build the grid a row for every hhpersonid and purpose.
new_grid <- expand_grid(
  hhpersonid = unique(new_trips$hhpersonid), trippurp = unique(new_trips$trippurp)
  ) %>%
  filter(trippurp != "-9")


# combine the grid and the trips
joined_grid_trips <- new_trips %>% 
  right_join(new_grid) %>%
  # if no trips were taken show they took 0 trips
  mutate(n = ifelse(is.na(n), 0, n)) %>%
  left_join(persons_edited, "hhpersonid") 
```

One major indication of mobility behavior is trip purpose. In the plot below, I show the counts (as a distribution) for number of trips in a day by trip purpose. The results show that travel behavior is different between the able bodied population and the wheelchair-dependent population. The major outlier is the home-based work trips, and that can be attributed to the major difference in employment among the ability groups.


```{r trip.purpose, message=F, warning=F, fig.width=8, fig.height=6, fig.align = "center"}
  # build distribution histogram from the built out grid of trips counted
  ggplot(joined_grid_trips, aes(x = n)) + 
  geom_histogram(
    aes(y = stat(width*density), fill = Ability), 
    binwidth = 0.8, colour = "black", size = .1, position = "dodge"
    ) +
  ggtitle("Comparing the Number of Trips in a Single Day", 
          subtitle = "Comparison of Ability") +
    labs(x = "Number of Trips", y = "Percent of Subgroup Population") +
    scale_fill_brewer(palette = "PuBuGn", direction = 1) + 
    theme(axis.text.x = 
            element_text(size  = 10, angle = 45, hjust = 1, vjust = 1)) +
    scale_x_continuous(breaks = c(0:10)) +
    scale_y_continuous(labels = percent_format()) +
    facet_wrap(~trippurp)
```

As the plot above only shows the Ability variable and trip purpose, it is likely that other variables have an influence on the difference in number of trips taken. Income, for example, may explain why wheelchair users travel differently than the abled population. 
This plot shows the same distribution of number of trips by purpose, but includes and compares the income groups. (Zoom in if necessary)

```{r purpose.grid, fig.width=12, fig.height=12}
ggplot(joined_grid_trips, aes(x = n)) + 
  geom_histogram(aes(y = stat(width*density), fill = Ability), binwidth = 0.8, colour = "black", size = .1, position = "dodge") +
    ggtitle("Comparing the number of trips by trip purpose and Income group", subtitle = "Comparison of Ability") +
    labs(x = "Number of Trips", y = "Percent of subgroup population") +
    scale_fill_brewer(palette = "PuBuGn", direction = 1) + 
    theme(axis.text.x = element_text(size  = 10, 
                                     angle = 45,
                                     hjust = 1,
                                     vjust = 1)) +
    scale_x_continuous(breaks = c(0:10)) +
    scale_y_continuous(labels = percent_format()) +
    facet_grid(trippurp ~ Income, scales = "free_x")
```

Note the changes in number of trips for wheelchair users as income increases. For home-based work trips, the percent of users who make zero trips decreases, while the number of zero trips for HBO and HBSHOP increase with income. 

```{r purpose.over.time, eval=F}
# make hhfaminc a factor in the joined_grid_trips data frame
joined_grid_trips <- joined_grid_trips %>%
  mutate(hhfaminc = as.factor(hhfaminc))

# plot the changes in number of trips over time for each purpose, 
# income group, and ability.
ggplot(data = joined_grid_trips, 
       aes(x = hhfaminc, y = stat(width*density))
       ) +
  geom_line(aes(color = Ability)) +
  geom_point(aes(color = Ability))
```

## Poisson Regression
It looks like "Ability" has no effect on trip purpose, instead, income may be what affects the number of trips by purpose the most. Future research will calculate the significance of Ability and Income on trip purpose counts.

```{r purpose.poisson, eval=FALSE}
joined_grid_trips$HBW <- ifelse(joined_grid_trips$trippurp == "HBW", TRUE, FALSE)
fit.hbw <- glm(HBW ~ Ability, Income, worker, family = "poisson", data = joined_grid_trips)
```











